Concurrent Programming in Harmony

### 2021-12-25 book/paper.tex
Pass over Ch 20

\texttt{Queue()} returns the initial value for a queue object
consisting of a <{None}> head and tail pointer and a lock.
The <{put(q, v)}> and <{get(q)}> methods both take a pointer <{q}> to the
queue object because both may modify the queue.
Before they access the value of the head or tail of the queue they first obtain
the lock.
When they are done, they release the lock.

@@

The <{synch}> library supports a synchronized message queue,
similar to the \texttt{Queue} object in Python.
Its interface is as follows:
\begin{itemize}
\item \texttt{Queue()} returns a new message queue;
\item <{put(q, item)}> adds <{item}> to the queue pointed to by <{q}>;
\item <{get(q)}> waits for and returns an item on the queue pointed to by <{q}>.
\end{itemize}

For those familiar with counting semaphores: note that a \texttt{Queue}
behaves much like a zero-initialized counting semaphore.
<{put}> is much like <{V}>, except that it is accompanied by data.
<{get}> is much like <{P}>, except that it also returns data.
Thus, synchronized queues can be considered a generalization of
counting semaphores.

\autoref{fig:actor} illustrates the actor approach.
There are three client threads that each want to be assigned
a unique identifier from the set $\{ 0, 1, 2 \}$.
Normally one would use a shared 0-initialized counter and a lock.
Each client would acquire the lock, get the value of the counter
and increment it, and release the lock.
Instead, in the actor approach the counter is managed by a
separate server thread.
The server never terminates, so it is spawned with the keyword
\textbf{eternal} to suppress non-terminating state warnings.
Each client sends a request to the server, consisting in this case
of simply the queue to which the server must send the response.
The server maintains a local, zero-initialized counter variable.
Upon receiving a request, it returns a response with the value of
the counter and increments the counter.  No lock is required.
The code is tested using the <{done}> array.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
<{tas(lk)}> & test-and-set on <{!lk}> \\
\hline
<{cas}(ptr, old, new)}> & compare-and-swap on <{!ptr}> \\
\hline
<{BinSem(v)}> & return a binary semaphore initialized to <{v}> \\
\hline
<{Lock()}> & return a binary semaphore initialized to <{False}> \\
\hline
<{acquire(bs)}> & acquire binary semaphore \texttt{!}<{bs}> \\
\hline
<{release(bs)}> & release binary semaphore \texttt{!}<{bs}> \\
\hline
<{Condition()}> & return a condition variable \\
\hline
<{wait(c, lk)}> & wait on condition variable \texttt{!}<{c}>
and lock <{lk}> \\ \hline
<{notify(c)}> & notify a thread waiting on condition variable \texttt{!}<{c}> \\
\hline
<{notifyAll(c)}> & notify all threads waiting on condition variable \texttt{!}<{c}> \\
\hline
<{Semaphore(cnt)}> & return a counting semaphore initialized to <{cnt}> \\
\hline
<{P(sema)}> & procure \texttt{!}<{sema}>  \\
\hline
<{V(sema)}> & vacate \texttt{!}<{sema}>  \\
\hline
\texttt{Queue}() & return a synchronized queue object \\
\hline
<{get(q)}> & return next element of <{q}>, blocking if empty \\
\hline
<{put(q, item)}> & add <{item}> to <{a}> \\
\hline
\end{tabular}

@@

\index{logical timestamp}%
Harmony provides support for testing linearizability,
but requires that the programmer
identifies what are known as \emph{linearization points}
in the implementation that indicate exactly \emph{which} sequential
execution the concurrent execution must align with.
\autoref{fig:queuelin} is a copy of \autoref{fig:queueconc} extended with
linearization points.
For each operation (<{get}> and <{put}>), the corresponding
linearization point must occur somewhere between acquiring and releasing
the lock.
%
Each linearization point execution is assigned a logical timestamp.
Logical timestamps are numbered $0, 1, ...$
To do so, we have added a counter (\texttt{time}) to the \texttt{Queue}.
Method \texttt{\_linpoint} saves the current counter in
\textbf{this}.\texttt{qtime} and increments the counter.
The \textbf{this} dictionary maintains \emph{thread-local state} associated
with the thread (\autoref{ch:harmonymachine})---it contains variables that
can be accessed by any method in the thread.

>>>>

<{Queue()}> returns the initial value for a queue object
consisting of a <{None}> head and tail pointer and a lock.
The <{put(q, v)}> and <{get(q)}> methods both take a pointer <{q}> to the
queue object because both may modify the queue.
Before they access the value of the head or tail of the queue they first obtain
the lock.
When they are done, they release the lock.

@@

The <{synch}> library supports a synchronized message queue,
similar to the <{Queue}> object in Python.
Its interface is as follows:
\begin{itemize}
\item <{Queue()}> returns an empty queue;
\item <{put(q, item)}> adds <{item}> to the queue pointed to by <{q}>;
\item <{get(q)}> waits for and returns an item on the queue pointed to by <{q}>.
\end{itemize}

For those familiar with counting semaphores: note that a <{Queue}>
behaves much like a zero-initialized counting semaphore.
<{put}> is much like <{V}>, except that it is accompanied by data.
<{get}> is much like <{P}>, except that it also returns data.
Thus, synchronized queues can be considered a generalization of
counting semaphores.

\autoref{fig:actor} illustrates the actor approach.
There are three client threads that each want to be assigned
a unique identifier from the set $\{ 0, 1, 2 \}$.
Normally one would use a shared 0-initialized counter and a lock.
Each client would acquire the lock, get the value of the counter
and increment it, and release the lock.
Instead, in the actor approach the counter is managed by a
separate server thread.
The server never terminates, so it is spawned with the keyword
\textbf{eternal} to suppress non-terminating state warnings.
Each client sends a request to the server, consisting in this case
of simply the queue to which the server must send the response.
The server maintains a local, zero-initialized counter variable.
Upon receiving a request, it returns a response with the value of
the counter and increments the counter.  No lock is required.

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
<{tas(lk)}> & test-and-set on <{!lk}> \\
\hline
<{cas}(ptr, old, new)}> & compare-and-swap on <{!ptr}> \\
\hline
<{BinSem(v)}> & return a binary semaphore initialized to <{v}> \\
\hline
<{Lock()}> & return a binary semaphore initialized to <{False}> \\
\hline
<{acquire(bs)}> & acquire binary semaphore \texttt{!}<{bs}> \\
\hline
<{release(bs)}> & release binary semaphore \texttt{!}<{bs}> \\
\hline
<{Condition()}> & return a condition variable \\
\hline
<{wait(c, lk)}> & wait on condition variable \texttt{!}<{c}>
and lock <{lk}> \\ \hline
<{notify(c)}> & notify a thread waiting on condition variable \texttt{!}<{c}> \\
\hline
<{notifyAll(c)}> & notify all threads waiting on condition variable \texttt{!}<{c}> \\
\hline
<{Semaphore(cnt)}> & return a counting semaphore initialized to <{cnt}> \\
\hline
<{P(sema)}> & procure \texttt{!}<{sema}>  \\
\hline
<{V(sema)}> & vacate \texttt{!}<{sema}>  \\
\hline
<{Queue}>() & return a synchronized queue object \\
\hline
<{get(q)}> & return next element of <{q}>, blocking if empty \\
\hline
<{put(q, item)}> & add <{item}> to <{a}> \\
\hline
\end{tabular}

@@

\index{logical timestamp}%
Harmony provides support for testing linearizability,
but requires that the programmer
identifies what are known as \emph{linearization points}
in the implementation that indicate exactly \emph{which} sequential
execution the concurrent execution must align with.
\autoref{fig:queuelin} is a copy of \autoref{fig:queueconc} extended with
linearization points.
For each operation (<{get}> and <{put}>), the corresponding
linearization point must occur somewhere between acquiring and releasing
the lock.
%
Each linearization point execution is assigned a logical timestamp.
Logical timestamps are numbered $0, 1, ...$
To do so, we have added a counter (\texttt{time}) to the <{Queue}>.
Method \texttt{\_linpoint} saves the current counter in
\textbf{this}.\texttt{qtime} and increments the counter.
The \textbf{this} dictionary maintains \emph{thread-local state} associated
with the thread (\autoref{ch:harmonymachine})---it contains variables that
can be accessed by any method in the thread.
